Random number generation and applications on GPU’s in R?

### 2021-09-16 pkg/gpuRandom/inst/documents/paper1_2021_5/paper1.tex
add paper1 to inst
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/diseasemapping@2851 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

\title{Random number simulation and applications on GPU’s in R?}
\Plaintitle{}
\Shorttitle{}

@@

We show the usage of \fct{gpuRandom::rnorm} by generating a $512 \times 1024$ matrix named \code{normal_matrix} of double presicion Gaussian random numbers, and assess Normality by making a histogram (left panel of Figure \ref{fig2}) and a Q-Q plot (right panel of Figure \ref{fig2}) of the generated random numbers. The points in the Q-Q plot in Figure \ref{fig2} fall along the diagonal line, we have good evidence the data are from Gaussian distribution. We find doing a Q-Q plot can occupy a large amount of time in the process of producing an R markdown file, so as an incidental work, we make the function \fct{gpuRandom::qqnorm} that does the quantile computation for the normal distribution part (adapted from \fct{stat::qnorm}) on GPU, which speeds up the Q-Q plot greatly.
\begin{knitrout}
\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
\begin{verbatim}
streams <- CreateStreamsGpu(Nstreams =512*128, keepInitial=1)
normal_matrix<-gpuRandom::rnorm(c(1024,512), streams=streams, 
                                Nglobal=c(512,128), type="double")
\end{verbatim}
\end{kframe}
\end{knitrout}
\begin{knitrout}
\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
\begin{verbatim}
avector<-as.vector(as.matrix(normal_matrix))
\end{verbatim}
\end{kframe}
\end{knitrout}
\begin{figure}[H]
\centering
\begin{knitrout}
\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
\begin{verbatim}
hist(avector,breaks=40, main = "histogram")
gpuRandom::qqnorm(avector, Nglobal=c(128,64), main="Q-Q Plot")
\end{verbatim}
\end{kframe}
\includegraphics[width=0.45\textwidth]{figure/histAndQQ-1} 
\includegraphics[width=0.45\textwidth]{figure/histAndQQ-2} 
\end{knitrout}
\caption{\label{fig2} Histogram and Q-Q plot of the GPU-generated random numbers.}
\end{figure}
Let's compare the time of generating a large-size matrix of more than 26.8 million Gaussian random numbers of double precision using the \fct{stats::rnorm} and using the \fct{gpuRandom::rnorm}. Both process are not time-consuming, however, using \fct{gpuRandom::rnorm} with $128 \times 128$ work-items is about 10 times (the number may fluctuate a bit from executions) faster than using the \fct{stats::rnorm}. The difference in elapsed time becomes larger when matrix size goes larger.
\begin{knitrout}
\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
\begin{verbatim}
system.time(gpuRandom::rnorm(c(16384,16384), streams=streams, 
                             Nglobal=c(512,128), type="double"))
##    user  system elapsed 
##   0.112   0.001   1.279
system.time(matrix(stats::rnorm(16384*16384),16384,16384))
##    user  system elapsed 
##  11.962   0.898  12.838
\end{verbatim}
\end{kframe}
\end{knitrout}

@@

\end{knitrout}
\begin{knitrout}
\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
\begin{verbatim}
#using CPU
stats::fisher.test(month,simulate.p.value = TRUE,B=1e6)$p.value
## [1] 0.4035866
system.time(stats::fisher.test(month,simulate.p.value = TRUE,B=1e6))
##    user  system elapsed 
##   9.753   0.004   9.740
\end{verbatim}
\end{kframe}
\end{knitrout}
\fct{stats::fisher.test} takes about 9.8 seconds (time varies a little bit from executions) for one million simulations, has a p-value of 0.4035.
\begin{knitrout}
\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
\begin{verbatim}
#using GPU
almost.1 <- 1 + 64 * .Machine$double.eps
observed_m= -gpuRandom::logfactSum(month, c(16,16))/almost.1
month_GPU<-vclMatrix(month,type="integer")
system.time(gpuRandom::fisher.sim(month_GPU, 1e6, streams=streams, 
            type="double", returnStatistics=FALSE, Nglobal = Nglobal))
##    user  system elapsed 
##   0.680   0.008   2.291
\end{verbatim}
\end{kframe}
\end{knitrout}
\fct{gpuRandom::fisher.sim} takes about 2.3 seconds, compared to the elapsed time of \fct{gpuRandom::fisher.sim},  \fct{gpuRandom::fisher.sim} takes less than a quater of the time. 
\begin{knitrout}
\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
\begin{verbatim}
result_monthgpu <- gpuRandom::fisher.sim(month_GPU, 1e6, streams=streams,
                   type="double", returnStatistics=TRUE,  Nglobal = Nglobal)
result_monthgpu$simnum
## [1] 999424
result_monthgpu$counts
## [1] 403274
result_monthgpu$p.value
## [1] 0.403507
\end{verbatim}
\end{kframe}
\end{knitrout}
We requested one million simulations on GPU while the actual number of simulation is 999424, and got 403274 cases whose test statitsic is below the observed threshold, so the p-value from \fct{gpuRandom::fisher.sim} for this table is about 0.4035, which is close enough to the p-value from \fct{stats::fisher.test}: 0.4035.

@@

\begin{knitrout}
\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
\begin{verbatim}
#using CPU
fisher.test(week,simulate.p.value = TRUE,B=1e7)$p.value
## [1] 0.0001189
system.time(fisher.test(week,simulate.p.value = TRUE,B=1e7))
##    user  system elapsed 
##  60.412   0.088  60.395
\end{verbatim}
\end{kframe}
\end{knitrout}
The ``week'' table has a much smaller p-value: around 0.000119, which should require a larger number of simulations to get a more accurate p-value.
\begin{knitrout}
\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
\begin{verbatim}
#using GPU
observed_w= -gpuRandom::logfactSum(week, c(16,16))/almost.1
week_GPU<-gpuR::vclMatrix(week,type="integer")
system.time(gpuRandom::fisher.sim(week_GPU, 1e7, streams=streams,
            type="double", returnStatistics=FALSE,Nglobal = Nglobal))
##    user  system elapsed 
##   0.613   0.010  10.839
\end{verbatim}
\end{kframe}
\end{knitrout}
\fct{stats::fisher.test} takes 60.4 seconds, while \fct{fisher.sim} takes about 10.8 seconds, the elapsed time is decreased to about 20\% of the time taken on CPU. 
\begin{knitrout}
\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
\begin{verbatim}
result_weekgpu<-gpuRandom::fisher.sim(week_GPU, 1e7, streams=streams,
                type="double",returnStatistics=TRUE,Nglobal = Nglobal)
result_weekgpu$simnum
## [1] 9994240
result_weekgpu$counts
## [1] 1272
result_weekgpu$p.value
## [1] 0.0001273734
\end{verbatim}
\end{kframe}
\end{knitrout}
With about ten million simulations, we get 1272 cases and a p-value around 0.0001273. We summarized the results in Table \ref{tab:summary}.

>>>>

\title{Random number generation and applications on GPU’s in R?}
\Plaintitle{}
\Shorttitle{}

@@

We show the usage of \fct{gpuRandom::rnorm} by generating a $512 \times 1024$ matrix named \code{normal_matrix} of double presicion Gaussian random numbers, and assess Normality by making a histogram (left panel of Figure \ref{fig2}) and a Q-Q plot (right panel of Figure \ref{fig2}) of the generated random numbers. The points in the Q-Q plot in Figure \ref{fig2} fall along the diagonal line, we have good evidence the data are from Gaussian distribution. We find doing a Q-Q plot can occupy a large amount of time in the process of producing an R markdown file, so as an incidental work, we make the function \fct{gpuRandom::qqnorm} that does the quantile computation for the normal distribution part (adapted from \fct{stat::qnorm}) on GPU, which speeds up the Q-Q plot greatly.
\begin{knitrout}
\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
\begin{verbatim}
streams <- CreateStreamsGpu(Nstreams =512*128, keepInitial=1)
normal_matrix<-gpuRandom::rnorm(c(1024,512), streams=streams, 
                                Nglobal=c(512,128), type="double")
\end{verbatim}
\end{kframe}
\end{knitrout}
\begin{knitrout}
\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
\begin{verbatim}
avector<-as.vector(as.matrix(normal_matrix))
\end{verbatim}
\end{kframe}
\end{knitrout}
\begin{figure}[H]
\centering
\begin{knitrout}
\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
\begin{verbatim}
hist(avector,breaks=40, main = "histogram")
gpuRandom::qqnorm(avector, Nglobal=c(128,64), main="Q-Q Plot")
\end{verbatim}
\end{kframe}
\includegraphics[width=0.45\textwidth]{figure/histAndQQ-1} 
\includegraphics[width=0.45\textwidth]{figure/histAndQQ-2} 
\end{knitrout}
\caption{\label{fig2} Histogram and Q-Q plot of the GPU-generated random numbers.}
\end{figure}
Let's compare the time of generating a large-size matrix of more than 26.8 million Gaussian random numbers of double precision using the \fct{stats::rnorm} and using the \fct{gpuRandom::rnorm}. Both process are not time-consuming, however, using \fct{gpuRandom::rnorm} with $128 \times 128$ work-items is about 10 times (the number may fluctuate a bit from executions) faster than using the \fct{stats::rnorm}. The difference in elapsed time becomes larger when matrix size goes larger.
\begin{knitrout}
\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
\begin{verbatim}
system.time(gpuRandom::rnorm(c(16384,16384), streams=streams, 
                             Nglobal=c(512,128), type="double"))
##    user  system elapsed 
##   0.102   0.008   1.278
system.time(matrix(stats::rnorm(16384*16384),16384,16384))
##    user  system elapsed 
##  12.028   0.857  12.862
\end{verbatim}
\end{kframe}
\end{knitrout}

@@

\end{knitrout}
\begin{knitrout}
\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
\begin{verbatim}
#using CPU
stats::fisher.test(month,simulate.p.value = TRUE,B=1e6)$p.value
## [1] 0.4035866
system.time(stats::fisher.test(month,simulate.p.value = TRUE,B=1e6))
##    user  system elapsed 
##   9.753   0.004   9.740
\end{verbatim}
\end{kframe}
\end{knitrout}
\fct{stats::fisher.test} takes about 9.8 seconds (time varies a little bit from executions) for one million simulations, has a p-value of 0.4035.
\begin{knitrout}
\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
\begin{verbatim}
#using GPU
almost.1 <- 1 + 64 * .Machine$double.eps
observed_m= -gpuRandom::logfactSum(month, c(16,16))/almost.1
month_GPU<-vclMatrix(month,type="integer")
system.time(gpuRandom::fisher.sim(month_GPU, 1e6, streams=streams, 
            type="double", returnStatistics=FALSE, Nglobal = Nglobal))
##    user  system elapsed 
##   0.672   0.012   2.287
\end{verbatim}
\end{kframe}
\end{knitrout}
\fct{gpuRandom::fisher.sim} takes about 2.3 seconds, compared to the elapsed time of \fct{gpuRandom::fisher.sim},  \fct{gpuRandom::fisher.sim} takes less than a quater of the time. 
\begin{knitrout}
\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
\begin{verbatim}
result_monthgpu <- gpuRandom::fisher.sim(month_GPU, 1e6, streams=streams,
                   type="double", returnStatistics=TRUE,  Nglobal = Nglobal)
result_monthgpu$simnum
## [1] 999424
result_monthgpu$counts
## [1] 403274
result_monthgpu$p.value
## [1] 0.403507
\end{verbatim}
\end{kframe}
\end{knitrout}
We requested one million simulations on GPU while the actual number of simulation is 999424, and got 403274 cases whose test statitsic is below the observed threshold, so the p-value from \fct{gpuRandom::fisher.sim} for this table is about 0.4035, which is close enough to the p-value from \fct{stats::fisher.test}: 0.4035.

@@

\begin{knitrout}
\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
\begin{verbatim}
#using CPU
fisher.test(week,simulate.p.value = TRUE,B=1e7)$p.value
## [1] 0.0001189
system.time(fisher.test(week,simulate.p.value = TRUE,B=1e7))
##    user  system elapsed 
##  60.412   0.088  60.395
\end{verbatim}
\end{kframe}
\end{knitrout}
The ``week'' table has a much smaller p-value: around 0.000119, which should require a larger number of simulations to get a more accurate p-value.
\begin{knitrout}
\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
\begin{verbatim}
#using GPU
observed_w= -gpuRandom::logfactSum(week, c(16,16))/almost.1
week_GPU<-gpuR::vclMatrix(week,type="integer")
system.time(gpuRandom::fisher.sim(week_GPU, 1e7, streams=streams,
            type="double", returnStatistics=FALSE,Nglobal = Nglobal))
##    user  system elapsed 
##   0.598   0.017  10.831
\end{verbatim}
\end{kframe}
\end{knitrout}
\fct{stats::fisher.test} takes 60.4 seconds, while \fct{fisher.sim} takes about 10.8 seconds, the elapsed time is decreased to about 20\% of the time taken on CPU. 
\begin{knitrout}
\definecolor{shadecolor}{rgb}{1, 1, 1}\color{fgcolor}\begin{kframe}
\begin{verbatim}
result_weekgpu<-gpuRandom::fisher.sim(week_GPU, 1e7, streams=streams,
                type="double",returnStatistics=TRUE,Nglobal = Nglobal)
result_weekgpu$simnum
## [1] 9994240
result_weekgpu$counts
## [1] 1272
result_weekgpu$p.value
## [1] 0.0001273734
\end{verbatim}
\end{kframe}
\end{knitrout}
With about ten million simulations, we get 1272 cases and a p-value around 0.0001273. We summarized the results in Table \ref{tab:summary}.
